Clustering galaxies at 0.5 < $z$ < 1: an unsupervised approach

### 2016-05-07 main.tex
Applied Pauline's and Hossein's result. Result0.01 is grammatically corrected (~ish)

\title[SOM: classifying high $Z$ galaxies]{Clustering galaxies at 0.5 < $z$ < 1: an unsupervised approach}
\author{rahmani.sahar }
\date{\today}
\author[S.~Rahmani, H.~Teimoorinia and P.~Barmby]{S.~Rahmani$^{1}$\thanks{E-mail:
srahma49@uwo.ca}, H.~Teimoorinia$^{2}$, P.~Barmby$^{1}$\\
$^{1}$Department of Physics $\&$ Astronomy, Western University, London, ON N6A 3K7, Canada\\
$^{2}$Department of Physics $\&$ Astronomy, University of Victoria, Finnerty Road, Victoria, British Columbia, V8P 1A1, Canada}
\maketitle

@@

\begin{abstract} %Most of this abstract is just a filler
In this paper we introduce a new method of classifying high red shift galaxies, based on their SED.
We use unsupervised self organizing map to train and validate an artificial neural network which can categorize the spectral energy distribution (SED) of galaxies. 
We utilize the spectral model introduced by Kinney et al. as a training set and a sample of 142 galaxies with 0.5 < $z$ < 1 a validation set.
We obtain physical properties of the sample galaxies measured using {\em CIGALE} code to study and compare the properties of each cluster.
In this paper we introduce a new method of classifying high red shift galaxies, based on their SED.
We used an unsupervised self organizing map (SOM)method, to create a network based in SED models of galaxies and used sample of galaxies to test the networks. 
We classify galaxies in different groups and studied the correlation between properties of the galaxies.
we conclude that the SOM is a very useful and accurate method to be used in classifying galaxies base on their morphology.

@@

\import{sections0.0/}{intro0.0.tex}
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%DATA
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------

\import{sections0.0/}{data0.0.tex}

@@

%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%Summery
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
\section{SUMMARY}
\label{sec: summary}
1) We introduced the self organizing maps to be used to categorizing the SED of galaxies.
\newline
2) Using this method we not only can categorized the galaxies with more precision but also we can see the relative similarity or dissimilarity between each type of galaxies.  
\newline
3) We show that using SOM, we could recognize the galaxies have similarity to two or more morphological types of galaxies.
\newline
4) We trained networks using sample of SEDs and test them with sample of galaxies with SEDs in the similar badpasses.
\newline
5) We plot the properties of the galaxies using new classification and see the correlation between age and SSFR and stellar mass of the galaxies.
\newline
6) We show that the properties of the galaxies in each groups is in good agreement between their morphological types.
\newline
7) We show that SOM can help us to easily identified new type of the galaxies, in large surveys very easy. 
\newline
8) This method can be used to find outliers very fast.
\newline

>>>>

\title[SOM: classifying high $Z$ galaxies]{Clustering galaxies at 0.5 < $z$ < 1: an unsupervised approach} %PB160426: suggest "Clustering galaxy spectra at $0.5<z<1$: an unsupervised approach" -- I think if you say "clustering galaxies" people will think of spatial clustering
\author{rahmani.sahar }
\date{\today}
\author[S.~Rahmani, H.~Teimoorinia and P.~Barmby]{S.~Rahmani$^{1}$\thanks{E-mail:
srahma49@uwo.ca}, H.~Teimoorinia$^{2}$, P.~Barmby$^{1}$\\
$^{1}$Department of Physics $\&$ Astronomy, Western University, London, ON N6A 3K7, Canada\\
$^{2}$Department of Physics $\&$ Astronomy, University of Victoria, Finnerty Road, Victoria, British Columbia, V8P 1A1, Canada}
\maketitle

@@

\begin{abstract} %Most of this abstract is just a filler % PB160426: OK, no comments on it for now.
In this paper we introduce a new method of clustering high red shift galaxies, based on their SED.
We use unsupervised self organizing map to train and validate an artificial neural network which can categorize the spectral energy distribution (SED) of galaxies. 
We utilize the spectral model introduced by Kinney et al. as a training set and a sample of 142 galaxies with 0.5 < $z$ < 1 a validation set.
We used physical properties of the sample galaxies measured using {\em CIGALE} code to study and compare the properties of each cluster.
In this paper we used an unsupervised self organizing map (SOM)method, to create a network based in SED models of galaxies and used sample of galaxies to test the networks. 
We classify galaxies in different groups and studied the correlation between properties of the galaxies.
we conclude that the SOM is a very useful and accurate method to be used in classifying galaxies base on their morphology.

@@

\import{sections0.01/}{intro0.01.tex}
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%DATA
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------

\import{sections0.01/}{data0.01.tex}

@@

%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%Summery
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
%----------------------------------------------------------------------------------------
\section{SUMMARY}
\label{sec: summary} %PB160427: not so sure I like the numbering of every sentence, or starting almost all of them with "We". But this is a style thing that we can discuss later.
1) We introduced the self organizing maps to be used to categorizing the SED of galaxies.
\newline
2) Using this method we not only can categorized the galaxies with more precision but also we can see the relative similarity or dissimilarity between each type of galaxies.  
\newline
3) We show that using SOM, we could recognize the galaxies have similarity to two or more morphological types of galaxies.
\newline
4) We trained networks using sample of SEDs and test them with sample of galaxies with SEDs in the similar badpasses.
\newline
5) We plot the properties of the galaxies using new classification and see the correlation between age and SSFR and stellar mass of the galaxies.
\newline
6) We show that the properties of the galaxies in each groups is in good agreement between their morphological types.
\newline
7) We show that SOM can help us to easily identified new type of the galaxies, in large surveys very easy. 
\newline
8) This method can be used to find outliers very fast.
\newline
%PB160427: Um, I'm not so sure you really demonstrated numbers 7 and 8. Those are interesting and it could be that a little more work on this and addition to the results section would be a good idea.

### 2016-05-07 sections0.01/intro0.01.tex
Applied Pauline's and Hossein's result. Result0.01 is grammatically corrected (~ish)

Advances in imaging techniques, and photometry and spectroscopy devises help to have higher resolution data and more detailed SED. 
Thus, the variations between SEDs become more clear and the classification of galaxies became more complicated.
Many groups used data from nearby galaxies in order to composite a thorough template for categorizing the spectral type of the galaxies~\citep[e.g.][]{Kinney93}~\citep[][hereafter K96]{Kinney96}~\citep[][]{Bershady00}.%and more references
These templates are mostly restricted to wavelengths from near-IR to ultraviolet (UV), so they contain main features from physical properties of galaxies i.e. age, star formation rate (SFR), stellar mass, wide range of the stellar population, interstellar medium (ISM)'s absorption and emission lines, and extinction from ISM of the galaxies.

One of the main usage of the SED templates is finding the morphological type of high redshift galaxies. 
However, since not two galaxies, even with same morphology, are exactly the same, classifying the SED of the galaxies are a challenge.
Many fitting methods are developed and used to find the best matches of template for each SED.
$\chi^2$ minimizing method is the most commonly used method to do so. 
Artificial neural networks (ANNs), K-mean clustering, and principle component analysis are the other methods to cluster and classifying the morphological type of the galaxies based on their SED \citep[e.g.][]{Allen13,Ordov14,Shi15}.

%ANNs
ANNs are very powerful tools to used in data processing and pattern recognitions problems.
Their information process method inspired by the way neurons in human brains work.
An ANN contains many interconnected units (neurons) which process data and work together to solve problems.
It uses sets of trainings method to learn about a nonlinear and complex relations between input and output data and how to apply it on new sets of data.
Studies shows that ANNs provides better fitting than minimizing chi square and would be an alternative choice for fitting data \citep[e.g.][]{Marquez91,Moayed09}

%SOMs and its advantages over supervised methods
Kohonen Self organizing map (or self organizing map, SOM) is an unsupervised neural network for mapping and visualizing a complex and non linear high dimension data introduced by~\citep{Kohonen82}.
The SOM is a technique that shows a simple geometry relationship of a non-linear high dimension data on a map \citep{Kohonen98}.
%Both advantage and disadvantage of the unsupervised method is that there are no previous predictions about the outcome.
In supervised method, a neural network would be trained using input data based on desired outcome.
This method is very useful for classification of data with specific target values (i.e. classifying SED of galaxies from a SED template).
On the other hand, in unsupervised method there is no prediction of output data.
This method classifies data based on their underlying structures and hidden variables.
The unsupervised method is a very helpful method to knowledge discovery of your data, or when the underlying structure of data is not well established (i.e. producing a template of SED of galaxies) %eh

%SOM in Astronomy
The utilization of the SOM in astronomy dates back to 1990s. 
\citet[][]{Odewahn92}, \citet[][]{Hernandez94}, and \citet[][]{Murtagh95} were among the first studies in astronomy that used SOM for their studies.
From classifying quasars' spectra to star/galaxy classifications, from gamma-ray bursts clustering to classificationf of light curves, this method have been used in various aspects in astronomy \citep[e.g.][]{Maehoenen95, Miller96,Andreon00,Balastegui01,Rajaniemi02,Brett04,Scaringi09}.
\citet{Geach12} used COSMOS data to demonstrate two of main application of SOMs; object classification and clustering, and photometric redshift estimation. 
The later one were subject of many studies \citep[e.g.][]{Kind14a}
\citet{In12} introduced a new clustering tool which is based on SOM method to analysing large data of spectra.
They used $\sim 60000$ spectra from the Sloan Digital Sky Survey \citep[SDSS;][]{Abazajian09}
to test their tool, and created very large SOMs to analyse the type of spectra/objects.
They also generated SOMs from quasar spectra in order to find unusual types of quasars spectra. Later \citet{Meusinger16} used these SOMs and updated data from SDSS and other surveys and found a new class of quasars.
The other application of the SOM is to find outliers or errors in the data.
\citet{Fustes13} used this method to produced a package to classify results from GAIA surveys and find the outlier or errors in the instruments or pipelines.

%What is this paper about
We obtained SED of a 142 galaxies with 0.5 < $z$ < 1 from~\citet[][hereafter T12]{Hossein12} paper.
T12 classified SED of a sample of galaxies using supervised neural network method, based on the spectral model presented by K96.
They classified SED of the 105 out of 142 galaxies using K96 template. 
The SED of the 37 galaxies from their samples could not be matched with any of those in the template. 
They found that the unclassified galaxies can be matched with a combined spectrum of the K96 model. 
They also showed that there are tight correlations between physical properties of galaxies and this correlations might be different based on the types of the galaxies.

In this paper, we used data from K96 template to train an unsupervised neural network.
The network clustered the SEDs and mapped those clusters and their relative relations.
We compared our results with K96 galaxies classification.
Then we classify the SED of the sample of 142 galaxies using the constructed SOM.

 In Section $\S$~\ref{sec: data}, we present the data. We describe the SOM in Section $\S$~\ref{sec: method}. The results of the SED classifications and comparing them with previous studoes are shown in Section $\S$~\ref{sec: result}. In Section $\S$~\ref{sec: summary}, we present the summary of our conclusion and the future works in this subject.

>>>>

Advances in imaging techniques, and photometry and spectroscopy devises help to have higher resolution data and more detailed SEDs. 
Thus, the variations between SEDs become more clear and the classification of galaxies became more complicated.
Many groups used data from nearby galaxies in order to composite a thorough template for categorizing the spectral type of the galaxies~\citep[e.g.][]{Kinney93}~\citep[][hereafter K96]{Kinney96}~\citep[][]{Bershady00,Mannucci01}.
Based on the usage of the templates, these templates could be restricted to various wavelengths.
In case of near-IR to ultraviolet (UV) wavelengths (which is pick output of stars), SED contains information about main features of physical properties of galaxies i.e. age, star formation rate (SFR), stellar mass, wide range of the stellar population, some information on interstellar medium (ISM)'s absorption and emission lines, and extinction from ISM of the galaxies.

SEDs of galaxies can be used to derive physical properties of galaxies, and from those one can find information about morphological type  galaxies.
Since not two galaxies, even with the same morphology, are exactly the same, classifying the SED of the galaxies are a challenge.
Many fitting methods are developed and used to find the best matches of template for each SED.
$\chi^2$ minimizing method is the most commonly used method to do so. 
Artificial neural networks (ANNs), K-mean clustering, and principle component analysis are the other methods to cluster and classifying the morphological type of the galaxies based on their SED \citep[e.g.][]{Allen13,Ordov14,Shi15}.

%ANNs
ANNs are very powerful tools to used in data processing and pattern recognitions problems.
Their information process method were inspired by the way neurons in human brains work.
An ANN contains many interconnected units (nodes or neurons) which process data and work together to solve problems.
It uses sets of trainings method to learn about a non-linear and complex relations between input and output data and how to apply it on new sets of data.
Studies shows that ANNs provides better fitting than minimizing chi square and would be an alternative choice for fitting data \citep[e.g.][]{Marquez91,Moayed09}.
In cases that both methods show the same correlation, ANNs still are better method of fitting due their faster results in large data bases \citep[][]{Gulati97}.

Neural networks can be trained in two methods; supervised and unsupervised.
In supervised method, a neural network would be trained using input data based on desired outcome.
This method is very useful for classification of data with specific target values (i.e. any pattern recognition with known templates).
On the other hand, in unsupervised method there is no prediction of output data.
This method classifies data based on their underlying structures and hidden variables.
The unsupervised method is a very helpful method to knowledge discovery of your data, or when the underlying structure of data is not well established (i.e. producing a template of SED of galaxies).

Kohonen Self organizing map (or self organizing map, SOM) is an unsupervised neural network for mapping and visualizing a complex and non-linear high dimension data introduced by~\citep{Kohonen82}.
The SOM is a technique that shows a simple geometry relationship of a non-linear high dimension data on a map \citep{Kohonen98}.
%SOM in Astronomy
The utilization of the SOM in astronomy dates back to 1990s. 
\citet[][]{Odewahn92}, \citet[][]{Hernandez94}, and \citet[][]{Murtagh95} were among the first studies in astronomy that used SOM for their studies.
From classifying quasars' spectra to star/galaxy classifications, from gamma-ray bursts clustering to classificationf of light curves, this method have been used in various aspects in astronomy \citep[e.g.][]{Maehoenen95, Miller96,Andreon00,Balastegui01,Rajaniemi02,Brett04,Scaringi09}.
\citet{Geach12} used COSMOS data to demonstrate two of main application of SOMs; object classification and clustering, and photometric redshift estimation. 
The later one were subject of many studies \citep[e.g.][]{Kind14a}
\citet{In12} introduced a new clustering tool which is based on SOM method to analysing large data of spectra.
They used $\sim 60000$ spectra from the Sloan Digital Sky Survey \citep[SDSS;][]{Abazajian09}
to test their tool, and created very large SOMs to analyse the type of spectra/objects.
They also generated SOMs from quasar spectra in order to find unusual types of quasars spectra. Later \citet{Meusinger16} used these SOMs and updated data from SDSS and other surveys and found a new class of quasars.
The other application of the SOM is to find outliers or errors in the data.
\citet{Fustes13} produced a package based on SOM to classify spectra from GAIA survey that were previously classified as "unknown" by SDSS pipeline. The package recognizes an astronomical object from an artificial errors, and then classifies the object based on its spectra.

%What is this paper about
In order to compare supervised and unsupervised method, in this paper, we used data from K96 template to train unsupervised neural networks. 
The network clustered the SEDs and mapped those clusters and their relative relations.
We compared our results with K96 galaxies classification.
Then we classify the SED of a sample of galaxies using the constructed SOM.

We used SED of a 142 galaxies with 0.5 < $z$ < 1 from~\citet[][hereafter T12]{Hossein12} paper.
T12 classified SED of a sample of galaxies using supervised neural network method, based on the spectral model presented by K96.
With supervised method, they could only classify SED of the 105 out of 142 galaxies.
The SED of the 37 galaxies from their samples could not be matched with any of those in the K96 templates. 
In order to classify the remaining 37 galaxies, they combined spectrum of the K96 templates.
The latter shows that, not a single type of galaxies in K96 template could describe the SED of the 37 galaxies.
The unsupervised methods on the other hand, have the freedom of classifying object between known classes, which can be used in this data to find a best SED class for the 37 galaxies.
T12 also showed that there are tight correlations between physical properties of galaxies and this correlations might be different based on the types of the galaxies.

 In Section $\S$~\ref{sec: data}, we present the data that we used to train and test our networks. We describe the SOM in Section $\S$~\ref{sec: method}. The results of the SED classifications and comparing them with previous studies are shown in Section $\S$~\ref{sec: result}. In Section $\S$~\ref{sec: summary}, we present the summary of our results and the future works in this subject.

### 2016-05-07 sections0.01/method0.01.tex
Applied Pauline's and Hossein's result. Result0.01 is grammatically corrected (~ish)

 %Kohonen Self organizing map (or self organizing map, SOM) is an unsupervised neural network for mapping and visualizing a complex and non linear high dimension data introduced by~\citep{Kohonen82}. 
% The SOM is a technique that shows a simple geometry relationship of a non-linear high dimension data on a map \citep{Kohonen98}.
 The SOM is a clustering method which reduces the dimension of data to lower dimensions usually 1 or 2D while preserving topological features of the original data.
 Results of SOM contains nodes (usually hexagonal ones) that arranged in 1D or 2D arrays.

 Each nodes may contain one or more samples from input data and distance between nodes represents similarity or dissimilarity of underlying samples. 
 In the way that similar data are closer together in the array and the further nodes go from each other, the more dissimilarity appears between their samples.
 %Sahar_self_notes: maybe "some" is better instead of one or more
 Moreover,a weight vector ``\boldit{W}" with same dimension of the input data associates with each node which will be change during the process and has a key factor in a position of nodes in the map.

 \cite{Geach12} presented the application of the SOM and demonstrate the algorithm of the SOM in detail. In this section we are going to talk briefly about the algorithm of the SOM, how we create our maps and a test model which will help to interpret our results. 
 \subsubsection{Algorithm of SOM} 
 \label{sec: algorithm}
     Assuming we have a data set which contains vectors, \boldit{V} $\in \Re^n$ and we want to map them on S1 by S2 map. 
     We start by creating S1 $\times$ S2 empty neurons. 
     The initial arrangement of these neurons depends on a map's topology provided by user.
     Since the topology of the map does not have any affect on the final result, we chose hexagonal topology which is the default topology for SOMs.
     Then, we assign a random weight vector \boldit{W} $\in \Re^n$ to each node.
     The process of creating SOM, happens over series of $N$ iterations. 
     During each iteration the weight vectors might change according to the Kohonen learning rule (equation~\ref{equ: weight adj}). 
      In each iteration:
     \begin{enumerate}
        \item Choose a random vector from our data set.
        \item Calculate the euclidean distance for each node j as  $D_j^2= \sum_{i=0}^{i=n} (V_i - W_i)^2$, and find a neuron with ``$D_{j_{min}}$". This neuron is the winner node and is calling Best Matching Unit (BMU). 
        \item  Compute the radius of the neighbourhood of the BMU to find nodes within this radius. The weight vectors of these nodes will be affected in the next steps. This value is arbitrary and initially can be set to be as high as half of the SOM size and then it decades exponentially over each iteration:
        \begin{equation}
            r^t_{BMU} = r^0_{BMU}e^{(-t/\tau)}
        \end{equation}
        where $\tau$ is a decay constant and usually set to be the same as number of iterations, $N$. $r^0_{BMU}$ and $r^t_{BMU}$ is the radius of the neighbourhood at 0th and $t$th iteration, respectively. 
        \item Change the weight vectors of the BMU and all the nodes within r(t) as:
        \begin{equation}
            \label{equ: weight adj}
            w(t+1)=w(t)+L(t) \times R(t) \times(v(t)-w(t))
        \end{equation}
        where $L(t) = L_0 e^{(-t/\tau)}$ is the learning factor which prevents divergence of the SOM and $R(t)=exp(-\frac{D_j^2}{2r^t_{BMU}})$ is the influence rate. $R(t)$ determines how weight of nodes in the neighbourhood of BMU will change.
        \item  Repeat these steps for $N$ times.
     \end{enumerate}

\subsection{Creating SOM}
\label{sec: create_som}
     In order to create SOM, we used {\tiny MATLAB} neural network toolbox~\citep[NNT,][]{matlabtolbox}.
     SOM in {\tiny NNT} can be created by {\tiny newsom} or {\tiny selforgmap} library which both work in two phases. 
     Phase one is the ``ordering phase". 
     This phase starts with maximum neighbourhood distance, and initial high learning factor usually 0.9 which is provided by user. 
     The ordering phase continues for requested number of iterations and it continues till the learning factor reduces to tuning phase leaning factor and the neighbourhood distance reaches to the number set by the user.

     The second phase is the ``tuning phase".
     In this phase the neighbourhood distance is at its minimum, but learning factor decreases very slowly.
     This minimum neighbourhood distance and slowly decreasing the leaning factor helps to fine tune the topology results and causes the more stable SOM. 
     The number of iterations in this tuning phase most be much more than the number iterations in ordering phase, to allow the tuning happens slowly. 
     We chose number of epochs the tuning phase be 3 times more than number of epochs in the ordering phase.

     To show our results, we combine two of the {\tiny NNT}'s built-in plots. 
     A Hits map, which shows number of hits for each neurons, and a distances map, which shows the same neurons as the hits map and the distance between them. 
     %To combine these two maps, we wrote number of hits in each neurons in the neurons in a distance map.
     In the maps, the purple hexagonal shape areas represent the neurons.
     %The red lines between neurons, show the immediate neighbours.
     The distances in a distance map are showing by colours.
     The darker colour represents the larger distance between neurons; whilst the lighter colours means there is small distance between neurons.
     In hit maps, neurons with zero hit left empty.

    %Although size of the SOM could be anything \cite{Vesanto05} suggested that the total number of  $5\sqrt{n}$ neurons provides the most sufficient size.
     Size of SOM maps is arbitrary and there is no rules or restrictions on choosing the size of maps. 
     Users usually must choose the size of grids based on their data set and their usages of the results.
     For training networks using 12 galaxies from K96, we created maps with sizes from $1\times2$ to $50\times50$ to choose the sufficient size.
     varying grids of the maps helps us to monitor whether grouping galaxies in smaller maps is because they have similar SEDs or its because of the shortage of the neurons.
     Based on size of our data and our results, we chose max of the grid size to be 1$\times$22 and 12$\times$12 in 1D and 2D maps, respectively.

     Also for each grid we created different SOM with different, learning factors, neighbourhood distances, and iteration numbers to find the optimize result for our sample.
     Based on our data we created our final SOMs with following initial values: number of iteration in ordering phase = 1000; ordering phase learning factor = 0.9; tuning phase learning factor= 0.02; and tuning phase neighbourhood distance to be 1.
     %All the other parameters are default values in {\tiny newsom} library and cannot be changed by the users.

@@

 We created  a mock sample of 27 galaxies to illustrate the results of SOMs and how they can be analyzed, and to show how this method works.
 The mock sample contains two information from each galaxy; first type of galaxies: type 1 or type 2, and either they are high or low star forming galaxies.
 We generated a SOM with size of $10 \times 10$ using the same initial values mentioned in the Section ~\ref{sec: create_som}.

 Fig. ~\ref{fig: sample} shows the SOM of the mock sample. 
 The upper plot is a distance map. 
 The axis show the position of the neurons in a $10 \times 10$ network. 
 The purple hexagons are illustration of the neurons, and the colours of the area between neighbours are showing the differences between weight of each two neurons.
 In distance maps colour white between neurons represents the minimum value of the differences between weights of neurons and the colour black shows the maximum values. 
 The gray cycle colour between white and black shows values between minimum and maximum with darker colour means more differences.

 The lower plot in the Fig.~\ref{fig: sample} shows a hit map.
 In this map similar to the distance map the axis shows the position of the neurons where the hexagonal shapes are neurons.
 %White neurons show the empty neurons.
 The purple neurons with a number on them shows the number of galaxy in that neurons.
 Colour coverage of neurons is different and depends on the number of the hits on the sample.
 The neuron with the maximum number of hits is completely coloured while the empty neurons are white.

Using this method, as we predicted, we were able to divide the mock sample galaxies in 4 distinct groups: Type 1 with high SFR, type 1 with low SFR, type 2 with high SFR and type 2 with low SFR. 
The upper plot in the Fig.~\ref{fig: sample}, clearly shows this division.
The upper part of the plot are high star forming galaxies, while the lower part of it belongs to the low star forming galaxies.
The left part of the plots is where type 1 galaxies belong to and the right side is for type 2 galaxies.
Grey to black colours shows the border between regions.
The lower plot in Fig.~\ref{fig: sample} shows only 4 neurons out of 100 were occupied. 
8 galaxies are type 1 galaxies with high SFR, 7 are type 2 galaxies with high SFR, 6 are type 1 galaxies with low SFR and the other 6 are type 2 galaxies with low SFR. 
The reason of that there are only 4 filled neurons in the network is the simplicity of the input data, and the fact that each entry only could get either 0 or 1 as for type of galaxy and 0 or 0.5 as an indicator of SFR.
It shows us although each galaxy had enough space to occupy any of the neurons in the network, because of the similarity of values they stayed in only four groups.

As we show in the following sections, in the real world with the real data, we never have two galaxies with exactly the same information.
Therefore, if the network has high enough neurons, the input data would eventually separate from each other and cluster into the smaller and smaller groups. 
However, if the information of the input data are pretty similar to each other, the number of neurons are going to be much higher than number of input samples. 
In similar cases, one can decide that the amount of similarity or dissimilarity between the input samples.

>>>>

 The SOM is a clustering method which reduces the dimension of data to lower dimensions usually 1 or 2D while preserving topological features of the original data.
 Results of SOM contains nodes (neurons) that arranged in 1D or 2D arrays \citep{Kohonen98}.

 Each nodes may contain one or more samples from input data and distance between nodes represents similarity or dissimilarity of underlying samples. 
 In the way that similar data are closer together in the array and the further nodes go from each other, the more dissimilarity appears between their samples.
 %Sahar_self_notes: maybe "some" is better instead of one or more
 A weight vector ``\boldit{W}" with same dimension of the input data associates with each node which will be change during the process and has a key factor in a position of nodes in the map.

 \cite{Geach12} presented the application of the SOM and demonstrate the algorithm of the SOM in detail. In this section we are going to talk briefly about the algorithm of the SOM, how we create our maps and a test model which will help to interpret our results. 
 \subsubsection{Algorithm of SOM} 
 \label{sec: algorithm}
     Assuming we have a data set which contains vectors, \boldit{V} $\in \Re^n$, and we want to map them on S1 by S2 map. 
     We start by creating S1 $\times$ S2 empty neurons. 
     The initial arrangement of these neurons depends on a map's topology provided by user.
     In the case of the 1D maps, since each neurons have two immediate neighbours the topology of the map does not have any affect on the final result and we can choose any topology.
     However, in 2D maps, the shape of the neurons specifies the number of the immediate neighbours for each neurons and it is up to user that which shape is more suitable for their data.
     In this paper, we chose hexagonal topology which gives each neuron six neighbours, and causes more interactions between neurons.
     Then, a random weight vector, \boldit{W} $\in \Re^n$, will be assigned to each node.
     The process of creating SOM, happens over series of $N$ iterations. 
     During each iteration the weight vectors might change according to the Kohonen learning rule (equation~\ref{equ: weight adj}). 
      In each iteration SOM code:
     \begin{enumerate}
        \item Chooses a random vector from our data set ($V_i$).
        \item Calculates the euclidean distance for each node, $j$, as  $D_j^2= \sum_{i=0}^{i=n} (V_i - W_i)^2$, and finds a neuron with minimum $D_j$, (``$D_{j_{min}}$"). This neuron is the winner node and is calling Best Matching Unit (BMU). 
        \item  Computes the radius of the neighbourhood of the BMU to find nodes within this radius. The weight vectors of these nodes will be affected in the next steps. This value is arbitrary and initially can be set to be as high as half of the SOM size and then it decades exponentially over each iteration:
        \begin{equation}
            r^t_{BMU} = r^0_{BMU}e^{(-t/\tau)}
        \end{equation}
        where $\tau$ is a decay constant and usually set to be the same as number of iterations, $N$. $r^0_{BMU}$ and $r^t_{BMU}$ is the radius of the neighbourhood at 0th and $t$th iteration, respectively. 
        \item Changes the weight vectors of the BMU and all the nodes within r(t) as:
        \begin{equation}
            \label{equ: weight adj}
            w(t+1)=w(t)+L(t) \times R(t) \times(v(t)-w(t))
        \end{equation}
        where $L(t) = L_0 e^{(-t/\tau)}$ is the learning factor which prevents divergence of the SOM and $R(t)=\exp(-\frac{D_j^2}{2r^t_{BMU}})$ is the influence rate. $R(t)$ determines how weight of nodes in the neighbourhood of BMU will change.
        \item  Repeats these steps for $N$ times.
     \end{enumerate}

\subsection{Creating SOM}
\label{sec: create_som}
     In order to create SOM, we used {\tiny MATLAB} neural network toolbox~\citep[NNT,][]{matlabtolbox}.
     SOM in {\tiny NNT} can be created by {\tiny newsom} or {\tiny selforgmap} library which both work in two phases. 
     Phase one is the ``ordering phase". 
     This phase starts with maximum neighbourhood distance, and initial high learning factor usually 0.9 which is provided by user. 
     The ordering phase continues for requested number of iterations.
     During the iterations, the learning factor reduces to tuning phase leaning factor and the neighbourhood distance reaches to tuning phase neighbourhood distance which both of the numbers set by the user.
     In the ordering phase, the changes in the learning factor and the neighbourhood distance adjust with the number of the iterations, in the way that at the last iteration, these two values reach to the initial values for the second phase.

     The second phase is the ``tuning phase".
     In this phase the neighbourhood distance is at its minimum, but learning factor decreases very slowly.
     This minimum neighbourhood distance and slowly decreasing the leaning factor helps to fine tune the topology results and causes the more stable SOM. 
     The number of iterations in this tuning phase most be much more than the number iterations in ordering phase, to allow the tuning happens slowly. %cite kohenen book?!
     We chose number of epochs the tuning phase be 3 times more than number of epochs in the ordering phase.

     To show our results, we used two of the {\tiny NNT}'s built-in plots. 
     A Hits map, which shows number of the times each neurons become the winner (hits), and a distances map, which shows the same neurons as the hits map and the distance between them.
     In the maps, the purple hexagonal shape areas represent the neurons.
     The distances in a distance map are showing by the grey cycle colours.
     The darker colour represents the larger distance between neurons; whilst the lighter colours means there is small distance between neurons.
     In hit maps, neurons with zero hit left empty.

    Size of SOM maps is arbitrary and there is no rules or restrictions on choosing the size of maps. 
    Although \cite{Vesanto05} suggested that the total number of  $5\sqrt{n}$ neurons provides the most sufficient size, users usually choose the size of grids based on their data set and their usages of the results.

@@

 We created  a mock sample of 27 galaxies to show how this method works, illustrate the results of SOMs, and how they can be analyzed.
 The mock sample contains two information from each galaxy; first type of galaxies: type 1 or type 2, and either they are high or low star forming galaxies. 
 We generated a SOM with size of $10 \times 10$ using the same initial values mentioned in the Section ~\ref{sec: create_som}.

 Fig. ~\ref{fig: sample} shows the SOM of the mock sample. 
 The upper plot is a distance map. 
 The axis show the position of the neurons in a $10 \times 10$ network.
 The lower plot in the Fig.~\ref{fig: sample} shows a hit map.
 In this map similar to the distance map the axis shows the position of the neurons where the hexagonal shapes are neurons.
 The purple neurons with a number on them shows the number of galaxy in that neurons.
 Colour coverage of neurons is different and depends on the number of the hits on the sample.
 The neuron with the maximum number of hits is completely coloured while the empty neurons are white.

Using this method, as we predicted, we were able to divide the mock sample galaxies in 4 distinct groups: Type 1 with high SFR, type 1 with low SFR, type 2 with high SFR and type 2 with low SFR. 
The upper plot in the Fig.~\ref{fig: sample}, clearly shows this division.
In that plot, the upper part belongs to high star forming galaxies, while the lower part belongs to the low star forming galaxies.
The left part of the plots is where type 1 galaxies belong to and the right side is for type 2 galaxies.
Grey to black colours shows the border between regions.
The lower plot in Fig.~\ref{fig: sample} shows only 4 neurons out of 100 were occupied. 
8 galaxies are type 1 galaxies with high SFR, 7 are type 2 galaxies with high SFR, 6 are type 1 galaxies with low SFR and the other 6 are type 2 galaxies with low SFR. 
Since the input data are so simple, each entry only could get either 0 or 1 as type of galaxy and 0 or 0.5 as an indicator of  high or low SFR, only four of the neurons were filled.
%The reason of that there are only 4 filled neurons in the network is the simplicity of the input data, and the fact that each entry only could get either 0 or 1 as for type of galaxy and 0 or 0.5 as an indicator of SFR.
We can concludes that although each galaxy had enough space to occupy any of the neurons in the network, because of the similarity of values they stayed only in four groups.
This network is considered as a train network, and if there is any data set with similar entries, we can use this network to cluster the new data set.

As we show in the following sections, in the real world with the real data, we never have two galaxies with exactly the same information, we have more than 2 dimensions, and more galaxy types. 
Therefore, if the network has high enough neurons, the input data would eventually separate from each other and cluster into the smaller and smaller groups. 
However, if the information of the input data are pretty similar to each other, the number of neurons are going to be much higher than number of input samples, to be able to separate the groups from each other. 
In those cases, one can decide that the amount of similarity or dissimilarity between the input samples.
